---
title: "Time series analysis of malaria in pregnancy, using Wavelet
transform and ARIMA models "
author: "Oniyelu Dolapo"
date: "`r Sys.Date()`"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Author-generated code underpins the findings in the manuscript

### Loading library

```{r}
library(deSolve)
library(zoo)
library(xts) 
library(dplyr)
require(ggplot2)
library(dlm)
library(forecast)
library(readr) 
library(biwavelet)
library(wavelets)
library(fields)
library(tseries)
library(car)
```

Import data: Low burden States

```{r, echo=FALSE}
setwd("C:\\Users\\Flourish PC\\Documents\\LEGEND_PAPER-20241119T151557Z-001\\LEGEND_PAPER\\NewHotSpotData")

```

### Select Region

South Southern region (Cross Rivers State) Datapoints: 2299, 19 LGAs, 228 records per year

```{r, echo=TRUE}
tsCross <- read.csv("C:\\Users\\Flourish PC\\Documents\\LEGEND_PAPER-20241119T151557Z-001\\LEGEND_PAPER\\NewHotSpotData\\CrossRiver_LGA.csv") 
head(tsCross, 3)
```

Note: Total Datapoints perculiarity per region

South Southern region (Cross Rivers State) Datapoints: 2299, 19 LGAs, 228 records per year

South Eastern region (Enugu State) Datapoints: 2178, 18 LGAs, 216 records per year

South Western region (Ondo State) Datapoints: 2299, 19 LGAs, 228 records per year

North Western region (Kebbi State) Datapoints: 2662, 22 LGAs, 264 records per year

North Central region (Niger State) Datapoints: 3146, 26 LGAs, 312 records per year

North Eastern region (Yobe State) Datapoints: 2178, 18 LGAs, 216 records per year

Also Note: datapoints between 2015-2020 for instance Cross Rivers State =228 * 6years
# Preprocessing

Feature selection, select Major columns of analysis

```{r, echo=TRUE}
datam <-  tsCross$Confirmed.uncomplicated.Malaria
dataMIP <- tsCross$Malaria.confirmed.pregnant.women
summary(dataMIP)#check outliers
summary(datam) #check outliers and missing data
```

### Select features

Prepare dataset for arrangement in year and LGAs view

```{r, echo=TRUE}
LGAsort <-  tsCross$organisationunitname
yearSORT <- tsCross$periodname

arrang<- cbind.data.frame(LGAsort, yearSORT, datam, dataMIP)
head(arrang,3)
```

### Visualize General malaria cases

```{r, echo=TRUE}

plot(datam, col="red",main="CrossRiver State: Malaria cases ",
     type="l",lwd="0.5", xlab="Time (months)", ylab = "Malaria cases", xaxt="n")
#Cross
t<-axis(side = 1, at= c(seq(0,2299,228)), labels= c(2015,2016,2017,2018,2019,2020,2021,2022,2023, 2024,2025))

```

### Visualize confirmed malaria in pregnant women cases

```{r}
plot(dataMIP, col="orange",
     type="l",lwd="0.5", xlab="Time (months)", ylab = "Malaria cases", xaxt="n", main="Malaria in pregnancy")
# #Cross
t<-axis(side = 1, at= c(seq(0,2299,228)), labels= c(2015,2016,2017,2018,2019,2020,2021,2022,2023, 2024,2025))
```

Note: x-axis format for other regions

Always adjust x-axis units based on selected state datapoints

Enugu t\<-axis(side = 1, at= c(seq(0,3146,228)), labels= c(2015,2016,2017,2018,2019,2020,2021,2022,2023, 2024,2025))

Ondo t\<-axis(side = 1, at= c(seq(0,2299,228)), labels= c(2015,2016,2017,2018,2019,2020,2021,2022,2023, 2024,2025))

kebbi t\<-axis(side = 1, at= c(seq(0,2662,264)), labels= c(2015,2016,2017,2018,2019,2020,2021,2022,2023, 2024,2025))

Niger t\<-axis(side = 1, at= c(seq(0,3146,312)), labels= c(2015,2016,2017,2018,2019,2020,2021,2022,2023, 2024,2025))

Yobe t\<-axis(side = 1, at= c(seq(0,2178,264)), labels= c(2015,2016,2017,2018,2019,2020,2021,2022,2023, 2024,2025))

### To check Skewness of data distribution

Histogram General malaria cases

```{r}

hist(datam, breaks=30, col="steelblue", main="Histogram", xlab="Values")

```

Histogram of malaria in pregnancy cases

```{r}
hist(dataMIP, breaks=30, col="steelblue", main="Histogram", xlab="Values")
```

Shapiro-Wilk Test for MIP cases

```{r}
shapiro_test <- shapiro.test(dataMIP)
shapiro_test
```

###Normalize Data: Data is skewed, thus we apply square root transformation

```{r}
or<-datam
orp<-dataMIP
datam <- sqrt(datam)
datam<-c(datam)
dataMIP<-sqrt(dataMIP)
dataMIP<-c(dataMIP)


```

## Fill in missing values using Kalman smoothening

Define the date sequence (yearSORT) and apply filters Keep only the first 19 days of each month (matching LGA records)

```{r}
yearSORT <- seq(as.Date("2015-01-01"), as.Date("2025-01-19"), by = "day")
yearSORT <- yearSORT[as.numeric(format(yearSORT, "%d")) <=19]  # Represents Cross RiversState 19 LGAs per month
```

Combine data into a data frame

```{r}
arrang1 <- cbind.data.frame(LGAsort, yearSORT, datam, dataMIP)
head(arrang1)

```

Arrange dataset by LGAsort to ensure fill is location inclined

```{r}
df <- as.data.frame(arrang1)
sort <- df %>% arrange(LGAsort)
head(sort)
```

Extract sorted columns for analysis

```{r}
datam <- sort$datam
dataMIP <- sort$dataMIP
LGAarrang <- sort$LGAsort
yeararrang <- sort$yearSORT
```

### Define the Kalman state-space model (Dynamic dW for pandemic effects)

```{r}
build_model <- function(disruption_effect) {
  dlmModPoly(order = 1, dV = 0.1, dW = mean(disruption_effect))  # Time-varying dW
}
```

Create pandemic disruption effect variable (for dataMIP smoothing)

```{r}
sort$disruption <- ifelse(
  yearSORT >= as.Date("2020-03-01") & yearSORT <= as.Date("2020-12-19"), 1,
  ifelse(yearSORT >= as.Date("2021-01-01"), exp(-0.1 * (as.numeric(yearSORT - as.Date("2021-01-01")))), 0)
)  # Exponential decay in recovery phases

```

Replace missing values in 'dataMIP' using Kalman filter and smoother

```{r}
categories <- unique(sort$LGAsort)  # Get unique LGAsort categories

for (LGAsort in categories) {
  # Subset data for the current category
  subset <- sort[sort$LGAsort == LGAsort, ]
  
  # Ensure subset is not empty
  if (nrow(subset) == 0) {
    print(paste("No data for LGAsort:", LGAsort))
    next
  }
  
  # Replace missing values for 'dataMIP' (full imputation)
  if ("dataMIP" %in% names(subset)) {
    dataMIP <- subset$dataMIP
    if (!is.null(dataMIP) && length(dataMIP) > 0) {
      model <- build_model(subset$disruption)  # Apply dynamic dW
      filtered_dataMIP <- dlmFilter(dataMIP, model)
      smoothed_dataMIP <- dlmSmooth(filtered_dataMIP)
      missing_indices <- which(is.na(dataMIP))
      dataMIP[missing_indices] <- smoothed_dataMIP$s[missing_indices]
      subset$dataMIP <- dataMIP
    }
  }
  
  # Replace missing values for 'datam' 
  if ("datam" %in% names(subset)) {
    datam <- subset$datam
    if (!is.null(datam) && length(datam) > 0) {
      filtered_datam <- dlmFilter(datam, dlmModPoly(order = 1, dV = 0.1, dW = 0.01))  
      
      smoothed_datam <- dlmSmooth(filtered_datam)
      missing_indices <- which(is.na(datam))
      datam[missing_indices] <- smoothed_datam$s[missing_indices] 
      subset$datam <- datam
    }
  }
  
    sort[sort$LGAsort == LGAsort, ] <- subset
}
head(sort)
```

Round results for final data consistency

```{r}
sort$datam <- sort$datam
sort$dataMIP <- sort$dataMIP

summary(sort)
```

Renaming for arranging back to monthly data, rather than based on LGAs

```{r}
sorts<- cbind.data.frame(sort)
#head(sorts)
```

Convert yearSORT into 'MMM-YY'

```{r}
sorts$yearSORT <- format(as.Date(sorts$yearSORT), "%b-%y")

# Define the custom order for yearsort levels
custom_order <- c("Jan-15", "Feb-15", "Mar-15", "Apr-15", "May-15", "Jun-15",
                  "Jul-15", "Aug-15", "Sep-15", "Oct-15", "Nov-15", "Dec-15",
                  "Jan-16", "Feb-16", "Mar-16", "Apr-16", "May-16", "Jun-16",
                  "Jul-16", "Aug-16", "Sep-16", "Oct-16", "Nov-16", "Dec-16",
                  "Jan-17", "Feb-17", "Mar-17", "Apr-17", "May-17", "Jun-17",
                  "Jul-17", "Aug-17", "Sep-17", "Oct-17", "Nov-17", "Dec-17",
                  "Jan-18", "Feb-18", "Mar-18", "Apr-18", "May-18", "Jun-18",
                  "Jul-18", "Aug-18", "Sep-18", "Oct-18", "Nov-18", "Dec-18",
                  "Jan-19", "Feb-19", "Mar-19", "Apr-19", "May-19", "Jun-19",
                  "Jul-19", "Aug-19", "Sep-19", "Oct-19", "Nov-19", "Dec-19",
                  "Jan-20", "Feb-20", "Mar-20", "Apr-20", "May-20", "Jun-20",
                  "Jul-20", "Aug-20", "Sep-20", "Oct-20", "Nov-20", "Dec-20",
                  "Jan-21", "Feb-21", "Mar-21", "Apr-21", "May-21", "Jun-21",
                  "Jul-21", "Aug-21", "Sep-21", "Oct-21", "Nov-21", "Dec-21",
                  "Jan-22", "Feb-22", "Mar-22", "Apr-22", "May-22", "Jun-22",
                  "Jul-22", "Aug-22", "Sep-22", "Oct-22", "Nov-22", "Dec-22",
                  "Jan-23", "Feb-23", "Mar-23", "Apr-23", "May-23", "Jun-23",
                  "Jul-23", "Aug-23", "Sep-23", "Oct-23", "Nov-23", "Dec-23",
                  "Jan-24", "Feb-24", "Mar-24", "Apr-24", "May-24", "Jun-24",
                  "Jul-24", "Aug-24", "Sep-24", "Oct-24", "Nov-24", "Dec-24", "Jan-25")

```

Apply custom order to yearSORT

```{r}
sorts$yearSORT <- factor(sorts$yearSORT, levels = custom_order, ordered = TRUE)
```

Ensure dataset remains sorted correctly

```{r}
sortss <- sorts %>% arrange(yearSORT)
head(sortss)
 #Plot original vs transformed data
hist(orp, main = "CrossRiver State MIP Incidence Original Data", xlab = "Value", col = "lightblue")
hist(sortss$dataMIP, main = "CrossRiver State MIP Incidence after Transformation", xlab = "Value", col = "lightblue")
write_csv(sortss, "Cross.csv")
Cross<-read.csv("Cross.csv")
```

# Prepare data for Pre-analysis and visualization of MIP and MC long-term trend

```{r}
datam<-Cross$datam
dataMIP<-Cross$dataMIP
LGAsort<-Cross$LGAsort
yearSORT<-Cross$yearSORT
disruption<-Cross$disruption
head(Cross)
```

Visualise preprocessed data - General cases

```{r}

plot(datam,  col="red", xaxt="n", main="CrossRiver State preprocessed Times series \n Malaria cases",
     type="l", xlab="Time (months)", ylab = "Malaria cases", lwd="0.5"     )
t<-axis(side = 1, at= c(seq(0,2299,228)), labels= c(2015,2016,2017,2018,2019,2020,2021,2022,2023, 2024,2025))
```

-MIP cases

```{r}
plot(dataMIP, col="orange",
     type="l",lwd="2", xlab="Time (months)", ylab = "Malaria cases", main="Malaria in pregnancy")
t<-axis(side = 1, at= c(seq(0,2299,228)), labels= c(2015,2016,2017,2018,2019,2020,2021,2022,2023, 2024,2025))
 
```

## Prepare time series for trend and seasonality decomposition

```{r}
datam=ts(datam, start=c(2015, 1),
         frequency=228)
head(datam)
 
```

Revert to MIP data Original available MIP data length for analysis (2015 - 2020)

```{r}
arrang<- cbind.data.frame(yearSORT, dataMIP)
arrang <- head(arrang, 1368)
head(arrang)
length(arrang$yearSORT)
length(arrang$dataMIP)
dataMIP <- arrang$dataMIP
dataMIP=ts(dataMIP, start=c(2015, 1),
           frequency=228)
length(dataMIP)
 
```

## Decompose timeseries to understand its structure, trends, and seasonality

```{r}

#decomposing trends
decomposedRes <- decompose(datam, type="additive") # use type = "additive" for additive components
decomposedRe <- decompose(dataMIP, type="additive") # 
 
```

Visualise decomposed MC

```{r}
plot (decomposedRes) # see plot below
 
```

Visualise decomposed MIP

```{r}
plot (decomposedRe) # see plot below
 
```

To get MC seasonal monthly trend in state

```{r}
month <- 1:12
Seasonwv<-decomposedRes$figure
Seasonw<-decomposedRe$figure

plot(Seasonwv, type='l', xlab = "Months",main = "CrossRiver State General Malaria Seasonal Trend", xaxt="n", ylab="Seasonality Index", col='red', las =2)
axis(side = 1, at = seq(1, 228,19), labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"))

 
```

To get MIP seasonal monthly trend in state

```{r}
plot(Seasonw, type='l', xlab = "Months",main = "CrossRiver State MIP Seasonal Trend", xaxt="n", ylab="Seasonality Index", col='red', las =2)
axis(side = 1, at = seq(1, 228,19), labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"))
```

# Variability of MIP cases over time compared to general MC cases

## Wavelet Coherence

```{r}
dataMIP<-arrang$dataMIP
length(datam)
length(dataMIP)
```

Adjust MC data to be same length as MIP (range between 2015-2020)

```{r}

############ 
######CONFIRM LENGTH OF DATA: 2015-2020
datam <- head(datam, 1368)
length(datam)
arrang<- cbind.data.frame(arrang,datam)
head(arrang)
 
```

format data for analysis and plot coherence

```{r}
cA <- ts(datam, start = c(2015, 1), frequency = 228)
cB <- ts(dataMIP, start = c(2015, 1), frequency = 228)
length(cB)
length(cA)

# Convert to numeric vectors
y1 <- as.numeric(unlist(cB))
y2 <- as.numeric(unlist(cA))
# Analyze coherency
A <- cbind(1:length(y1), y1)
B <- cbind(1:length(y2), y2)
#no. of simulations
########
nrands =50
wtc.Cross = wtc(A,B, nrands =nrands)
# Set margins
par(oma=c(0,0,0,1 ), mar =c(5,4,5,5) +0.1)
coherence<-plot(wtc.Cross,     plot.phase = TRUE,xaxt = "n",
                lty.coi = 1, col.coi = "purple", lwd.coi=2,
                lwd.sig = 2, arrow.lwd = 0.03, arrow.len=0.08,
                ylab="Frequency", xlab = "Time(Months)",
                plot.cb = T, main="CrossRiver State Coherence plot")

#cross
t<-axis(side = 1, at= c(seq(1,1368,228)), labels= c(2015,2016,2017,2018,2019,2020))
 
```

#Forecast of future trends of MIP cases \## Denoising data for wavelet-ARIMA/SARIMAX modelling

```{r}
head(cB) 
y <- as.numeric(unlist(cB))
length(cB)
# Parameters
omega <- 6  # Angular frequency
# Morlet wavelet function
morlet_wavelet <- function(t, scale, omega, A) {
  A * exp(1i * omega * t) * exp(-t^2 / (2 * scale^2))
}
# Reconstruct the signal using wavelet coefficients
reconstructed_signal <- numeric(length(y))
threshold <- 0.5  # Example threshold value
for (j in seq(0, 3)) {  # Adjusted to 3 levels as an example
  scale <- 2^(j / 2)
  A <- 1 / scale  # Amplitude inversely proportional to scale
  for (k in 1:length(y)) {
    t <- (1:length(y)) - k
    wt <- sum(morlet_wavelet(t, scale, omega, A) * y)  # Wavelet transform
    if (abs(wt) >= threshold) {  # Apply thresholding
      reconstructed_signal[k] <- reconstructed_signal[k] +
        Re(wt) / sqrt(scale)
    }
  }
}

```

Visualise MIP before denoise

```{r}
plot(y, type = "l", xaxt="n", col = "red", main = " Preprocessed CrossRiver State MIP Time series", ylab = "Malaria Cases")
t<- axis(side = 1, at= c(seq(1,length(y),228)), labels= c(2015,2016,2017,2018,2019,2020))
```

Visualise denoised MIP data

```{r}
plot(cB,  type = "l", col = "purple", main = "Reconstructed CrossRiver State MIP Time series", ylab = "Malaria Cases")
t<- axis(side = 1, at= c(seq(1,length(y),228)), labels= c(2015,2016,2017,2018,2019,2020))
###

# Save the reconstructed signal as a CSV file
MIPCross <- reconstructed_signal
mDNCross <- as.data.frame(MIPCross)
write_csv(mDNCross, "mDNCross.csv")
```

###use denoised data to set initial value for ARIMA/ SARIMAX model

```{r}
cB<- read.csv("mDNCross.csv")
cB=ts(cB, start=c(2015, 1),
      frequency=228)
length(cB)
```

### Split the data into training and test sets

```{r}
train_size <- floor(0.70 * length(cB))
train_cB <- cB[1:train_size]
test_cB <- cB[(length(train_cB) + 1):length(cB)]
```

### Plot ACF and PACF

```{r}
acf(train_cB, main = "CrossRiver State: \n ACF of Training Data")
pacf(train_cB, main = "PACF of Training Data")
```

###Perform ADF test to determine the number of differences needed

```{r}
# 

adf_test <- adf.test(train_cB)
adf_test
station<-if (adf_test$p.value > 0.05) {
  train_cB_diff <- diff(train_cB)
  adf_test_diff <- adf.test(train_cB_diff)
  d <- 1
  if (adf_test_diff$p.value > 0.05) {
    train_cB_diff2 <- diff(train_cB_diff)
    adf_test_diff2 <- adf.test(train_cB_diff2)
    if (adf_test_diff2$p.value <= 0.05) {
      d <- 2
    }
  }
} else {
  d <- 0
}

d

p <- 1
q <- 3

# # Fit the ARIMA model with initial values
initial_model <- arima(train_cB, order = c(p, d, q))

# Print the initial model summary
summary(initial_model)
AIC(initial_model)
```

# ARIMA analysis

```{r}

find_best_arima <- function(train_cB, test_cB) {
  best_aic <- Inf
  best_order <- c(0, 0, 0)
  best_model <- NULL
  last_three_models <- list()
  best_models <- list()  # Added to store models
  metrics <- data.frame(Model = character(), AIC = numeric(), BIC = numeric(), RMSE = numeric(), MAPE = numeric(), VIF = numeric(), stringsAsFactors = FALSE)
  
# Determine initial values for p, d, q (example values, update as needed)
for (p in 1:3) {
    for (d in 0:2) {
      for (q in 0:3) {
        model <- tryCatch({
          arima(train_cB, order = c(p, d, q), optim.method = "BFGS", optim.control = list(maxit = 1000))
        }, error = function(e) NULL)
        
        if (!is.null(model)) {
          aic <- AIC(model)
          bic <- BIC(model)
          forecast_values <- forecast(model, h = length(test_cB))
          
          # Calculate RMSE and MAPE using the test data
          rmse <- sqrt(mean((forecast_values$mean - test_cB)^2, na.rm = TRUE))
          mape <- mean(abs((forecast_values$mean - test_cB) / test_cB), na.rm = TRUE) * 100
          
          # Calculate VIF
          vif_values <- tryCatch({
            vif(model)
          }, error = function(e) rep(NA, length(coef(model))))
          
          avg_vif <- mean(vif_values, na.rm = TRUE)
          
          metrics <- rbind(metrics, data.frame(Model = paste(p, d, q, sep = ","), AIC = aic, BIC = bic, RMSE = rmse, MAPE = mape, VIF = avg_vif))
          
          # Store the best models
          best_models[[paste(p, d, q, sep = ",")]] <- model
          
          if (aic < best_aic && !is.nan(aic)) {
            best_aic <- aic
            best_order <- c(p, d, q)
            best_model <- model
          }
          
          # Store the last three models
          last_three_models <- append(last_three_models, list(model))
          if (length(last_three_models) > 3) {
            last_three_models <- last_three_models[-1]
          }
        }
      }
    }
  }
  
  return(list(best_model = best_model, best_order = best_order, best_aic = best_aic, last_three_models = last_three_models, metrics = metrics, best_models = best_models))
}

# Find the best ARIMA model
best_arima <- find_best_arima(train_cB, test_cB)
best_arima
```

# Print and Sort to extract the metrics for all models

```{r}

# Print the metrics for all models
print(best_arima$metrics)
best_model_metrics <- best_arima$metrics[which.min(best_arima$metrics$AIC), ]
best_model_metrics

# Sort the metrics data frame by AIC
sorted_metrics <- best_arima$metrics[order(best_arima$metrics$AIC), ]

# Extract the metrics for the best four models
top_four_metrics <- head(sorted_metrics, 4)
top <- rbind(top_four_metrics, best_model_metrics)
ARIMAmodels <- as.data.frame(top)
```

### Step 8: Generate Forecasts for Each of the Top 4 Models

```{r}
# Step 8: Generate Forecasts for Each of the Top 4 Models
forecasted_values <- list()
for (i in 1:4) {
  model_name <- top_four_metrics$Model[i]
  model <- best_arima$best_models[[model_name]]  # Corrected to reference stored models
  
  forecasted_values[[model_name]] <- tryCatch(
    forecast(model, h = length(test_cB))$mean,  # Removed undefined `xreg = test_disruption`
    error = function(e) return(rep(NA, length(test_cB)))  # Handle errors gracefully
  )
}

# Step 9: Format Forecast Results for Comparison
forecast_df <- as.data.frame(forecasted_values)
colnames(forecast_df) <- top_four_metrics$Model  # Assign model names as column headers
# Add test data for direct comparison
forecast_df$Actual_Test_Data <- test_cB

read.csv("Cross.csv")
# Save the forecasted values for the top 4 models in a CSV file
head(forecast_df)
n <- nrow(forecast_df) # Get the number of rows in y
yearSORT<- Cross$yearSORT
yr<-as.data.frame(tail(yearSORT, n)) # Extract the last n values from yearSORT
f<-cbind(yr, forecast_df)
head(f) 
write.csv(f, "ARIMAtop_4_testCross_forecasts.csv", row.names = FALSE)

write_csv(top, "CrossARIMA.csv")
head(top)
#CrossAri<-read.csv("CrossARIMA.csv")
read.csv("CrossARIMA.csv")
```

```{r}
acf_residuals <- residuals(best_arima$best_model)
acf(acf_residuals, main = "ACF of Residuals for Best ARIMA Model")
pacf_residuals <- residuals(best_arima$best_model)
pacf(pacf_residuals, main = "PACF of Residuals for Best ARIMA Model")

```

```{r}
checkresiduals(best_arima$best_model)
```

#ARIMA residuals testing applying the Jarque–Bera test for normality, Lagrange multiplier (LM) test for ARCH (Autoregressive conditional heteroskedasticity) effect and the Ljung–Box (LB) test for autocorrelation function

```{r}
#Perform the Jarque-Bera test on residuals
residuals <- acf_residuals
jb_test <- jarque.bera.test(residuals)
print(jb_test)
jb_stat <- jb_test$statistic
jb_p_value <- jb_test$p.value
cat("Jarque-Bera Test: Q-statistic =", jb_stat, ", p-value =", jb_p_value, "\n")
jb <- data.frame(jb_stat, jb_p_value)
library(lmtest)
library(FinTS)
# Perform the LM test for ARCH effect on residuals
lm_test <- ArchTest(residuals, lags = 10)
print(lm_test)
lm_stat <- lm_test$statistic
lm_p_value <- lm_test$p.value
cat("ARCH LM Test: Q-statistic =", lm_stat, ", p-value =", lm_p_value, "\n")
lm<- data.frame(lm_stat, lm_p_value)
# Perform the Ljung-Box test on residuals
Box_test <- Box.test(residuals, type = "Ljung-Box", lag = 10)
print(Box_test)
box_stat <- Box_test$statistic
box_p_value <- Box_test$p.value
cat("Ljung-Box Test: Q-statistic =", box_stat, ", p-value =", box_p_value, "\n")
bm<- data.frame(box_stat, box_p_value)

resiCross<- cbind("Cross",jb,lm,bm)
#resiUpdate <- rbind(resiUpdate, resiCross)
write.csv(resiCross, "resiCross.csv")
head(resiCross)

best_model <- best_arima$best_model

```

###FORECAST

```{r}
Cross<-read.csv("Cross.csv")
cB<- (Cross$dataMIP)^2
cB <- ts(cB, start = c(2015, 1), frequency = 228)
#Generate forecast values
#fore_cB<- length(1:(length(test_cB) + (228 * 2)))
cases_fit<-arima(x=(cB), order=c(3,1,3)
              #  ,seasonal=list(order=c(2,0,0))
              )
                 
forecast_values <- forecast(cases_fit, h = 228*2, level = c(95))
forecast_values


```

```{r}
# Plot historical values and forecast while maintaining correct limits
autoplot(cB) + 
  autolayer(forecast_values, series = "Forecast", PI = TRUE) +
  ggtitle("ARIMA(3,1,3) Forecast for CrossRiver State") +
  xlab("Time (months)") + 
  ylab("MIP Cases") +
  scale_x_continuous(breaks = seq(2015, 2027, 1)) +
  theme_minimal()
```

#SARIMAX analysis

###Accounting for exogenous factor (Pandemic disruption)

```{r}
cB <- read.csv("mDNCross.csv") #### already in use
cB = ts(cB, start = c(2015, 1), frequency = 228)

# Split the data into training and test sets
train_size <- floor(0.70 * length(cB))
train_cB <- cB[1:train_size]
test_cB <- cB[(length(train_cB) + 1):length(cB)]

disrupt <- head(disruption, 1368)
length(disrupt)
tail(disrupt)

train_disruption <- disrupt[1:train_size]  # Training regressor
test_disruption <- disrupt[(train_size + 1):length(cB)]  # Testing regressor

```

### Fit SARIMAX Models & Store AIC Values

```{r}
models <- list()
model_aic_values <- c()  # Store AIC values for sorting
model_bic_values <- c()  # Define BIC storage

#use denoising initial
for (p in 1:3) {
  for (d in 0:2) {
    for (q in 0:3) {
      for (P in 0:2) {
        for (D in 0:2) {
          for (Q in 0:2) {
            model <- tryCatch(
              Arima(train_cB, order = c(p, d, q), seasonal = c(P, D, Q), xreg = train_disruption),
              error = function(e) return(NULL)  # Catch errors
            )
            if (!is.null(model)) {
              model_name <- paste(p, d, q, P, D, Q)
              models[[model_name]] <- model
              model_aic_values[model_name] <- AIC(model)
              model_bic_values[model_name] <- BIC(model)  # Store BIC
            }
          }
        }
      }
    }
  }
}

#summary(models)
```

###Select Best 4 Models Based on AIC

```{r}
sorted_indices <- order(model_aic_values)  # Sorting based on AIC
print(model_bic_values )
print(model_aic_values)
print(data.frame(AIC = model_aic_values, BIC = model_bic_values))

```

Save Model Comparisons (AIC & BIC)

```{r}
best_model_metrics <- data.frame(
  Model = names(models),
  AIC = sapply(models, AIC),
  BIC = sapply(models, BIC)
)
summary(best_model_metrics)

```

Get different AIC values,Sort in ascending order and select

```{r, echo=FALSE}
unique_aic_values <- unique(best_model_metrics$AIC)  
sorted_aic_values <- sort(unique_aic_values) 
selected_models <- list()
for (aic_value in sorted_aic_values) {
  subset_models <- best_model_metrics[best_model_metrics$AIC == aic_value, ]  
  best_bic_model <- subset_models[which.min(subset_models$BIC), ]  
  
  selected_models[[as.character(aic_value)]] <- best_bic_model  
}
selected_models_df <- do.call(rbind, selected_models)
print(selected_models_df)

best_models <- models[selected_models_df$Model]  

```

###Evaluate Performance (RMSE & MAPE)

```{r, echo=FALSE}
evaluate_model <- function(model) {
  forecasted <- tryCatch(
    forecast(model, xreg = test_disruption, h = length(test_cB)),  # Forecast on test data
    error = function(e) return(NULL)  # Catch errors
  )
  # Handle models that fail to forecast properly
  if (is.null(forecasted)) return(c(AIC = AIC(model), BIC = BIC(model), RMSE = NA, MAPE = NA))
  # Compute RMSE and MAPE
  rmse <- sqrt(mean((test_cB - forecasted$mean)^2, na.rm = TRUE))  # Root Mean Squared Error
  mape <- mean(abs((test_cB - forecasted$mean) / test_cB), na.rm = TRUE) * 100  # Mean Absolute Percentage Error
  
  return(c(AIC = AIC(model), BIC = BIC(model), RMSE = rmse, MAPE = mape))
}


metrics <- lapply(best_models, evaluate_model)  # Evaluate performance
metrics_df <- do.call(rbind, metrics)  # Convert to data frame
rownames(metrics_df) <- names(best_models)  # Assign model names
write.csv(metrics_df, "model_evaluation_metrics.csv", row.names = TRUE)
length(metrics_df)
metrics_df <- as.data.frame(do.call(rbind, metrics))  # Ensure it's a data frame

```

Sort Models Based on RMSE

```{r, echo=FALSE}
sorted_models <- metrics_df[order(metrics_df$RMSE), ]  # Sort by lowest RMSE
top_3_models <- sorted_models[1:3, ]  # Get the top 3 models

forecasted_values <- list()

for (i in 1:3) {
  model_name <- rownames(top_3_models)[i]
  model <- best_models[[model_name]]
  
  forecasted_values[[model_name]] <- tryCatch(
    forecast(model, xreg = test_disruption, h = length(test_cB))$mean,  # Extract only forecasted values
    error = function(e) return(rep(NA, length(test_cB)))  # Handle errors gracefully
  )
}

#Format Forecast Results for Comparison
forecast_df <- as.data.frame(forecasted_values)
colnames(forecast_df) <- rownames(top_3_models)  # Assign model names as column headers

# Add test data for direct comparison
forecast_df$Actual_test_cB <- test_cB
# Save the forecasted values for the top 3 models in a CSV file
n <- nrow(forecast_df) # Get the number of rows in y
yearSORT<- Cross$yearSORT
yr<-as.data.frame(tail(yearSORT, n)) # Extract the last n values from yearSORT
f<-cbind(yr, forecast_df)
write.csv(f, "SARIMAtop_3_testCross_forecasts.csv", row.names = FALSE)
head(forecast_df)
```

Select the Final Best Model

```{r}
best_model_index <- which.min(top_3_models$RMSE)  # Find model with lowest RMSE
best_model_name <- rownames(top_3_models)[best_model_index]
best_model <- best_models[[best_model_name]]

summary(best_model)

```

Extract forecasted values for the best model

```{r}
best_model_forecast <- forecast(best_model, xreg = test_disruption, h = length(test_cB))$mean
best_model_forecast_df <- data.frame(
  Forecasted_Values = best_model_forecast,
  Actual_test_cB = test_cB
)


```

Save the best model’s forecasted values in another CSV file

```{r}
write.csv(best_model_forecast_df, "testCross_model_forecast.csv", row.names = FALSE)
head(best_model_forecast_df)

```

Extracting model parameters safely

```{r}
best_model_params_df <- data.frame(
  Model = best_model_name,  # Store the model name
  p = best_model$arma[1],  # AR Order
  d = best_model$arma[6],  # Differencing Order
  q = best_model$arma[2],  # MA Order
  P = best_model$arma[3],  # Seasonal AR Order
  D = best_model$arma[7],  # Seasonal Differencing Order
  Q = best_model$arma[4],  # Seasonal MA Order
  Intercept = ifelse(!is.null(coef(best_model)[1]), coef(best_model)[1], NA)  # Intercept term
)

# Print formatted best model parameters
print(best_model_params_df)

# Save best model parameters to a CSV file
write.csv(best_model_params_df, "CrossSARIMA_model_parameters.csv", row.names = FALSE)

```

###Extract residuals from the final best model

```{r}
acf_residuals <- residuals(best_model)
acf(acf_residuals, main = "ACF of Residuals for Best SARIMAX Model")
pacf_residuals <- residuals(best_model)
pacf(pacf_residuals, main = "PACF of Residuals for Best SARIMAX Model")

```

Perform residual diagnostic check

```{r}
checkresiduals(best_model)

```

Residuals testing applying the Jarque–Bera test for normality, Lagrange multiplier (LM) test for ARCH (Autoregressive conditional heteroskedasticity) effect and the Ljung–Box (LB) test for autocorrelation function

```{r, echo=FALSE}
#Perform the Jarque-Bera test on residuals
residuals <- acf_residuals
jb_test <- jarque.bera.test(residuals)
print(jb_test)
jb_stat <- jb_test$statistic
jb_p_value <- jb_test$p.value
cat("Jarque-Bera Test: Q-statistic =", jb_stat, ", p-value =", jb_p_value, "\n")
jb2 <- data.frame(jb_stat, jb_p_value)
library(lmtest)
library(FinTS)
# Perform the LM test for ARCH effect on residuals
lm_test <- ArchTest(residuals, lags = 10)
print(lm_test)
lm_stat <- lm_test$statistic
lm_p_value <- lm_test$p.value
cat("ARCH LM Test: Q-statistic =", lm_stat, ", p-value =", lm_p_value, "\n")
lm2<- data.frame(lm_stat, lm_p_value)
# Perform the Ljung-Box test on residuals
Box_test <- Box.test(residuals, type = "Ljung-Box", lag = 10)
print(Box_test)
box_stat <- Box_test$statistic
box_p_value <- Box_test$p.value
cat("Ljung-Box Test: Q-statistic =", box_stat, ", p-value =", box_p_value, "\n")
bm2<- data.frame(box_stat, box_p_value)

resiCross2<- cbind("Cross2",jb2,lm2,bm2)
#resiUpdate <- rbind(resiUpdate, resiCross)
write.csv(resiCross2, "resiCross2.csv")
head(resiCross2)

```

Prepare Exogenous Variables for Future Forecast, using original data unit

```{r}
Cross<-read.csv("Cross.csv")
dataMIP<-(Cross$dataMIP)^2
dataMIP_ts <- ts(dataMIP, start = c(2015, 1), frequency = 228)
disruption<-(Cross$disruption)^2
length(dataMIP)
summary(dataMIP)
dataMIP <- ts(dataMIP, start = c(2015, 1), frequency = 228)  # 228 points per year
yearSORT <- seq(as.Date("2015-01-01"), as.Date("2025-01-19"), by = "day") # Keep only the first 19 days of each month (matching LGA records) 
yearSORT <- yearSORT[as.numeric(format(yearSORT, "%d")) <=19] # Represents 19 LGAs per month
disruption <- ifelse( yearSORT >= as.Date("2020-03-01") & yearSORT <= as.Date("2020-12-19"), 1, 
                           ifelse(yearSORT >= as.Date("2021-01-01"), exp(-0.1 * (as.numeric(yearSORT - as.Date("2021-01-01")))), 0) )


```

###Fit Best SARIMAX Model with Exogenous Regressors

```{r}
# Step 1: 
cases_fit <- Arima(dataMIP, order = c(2,0,2), seasonal = c(0,0,0), xreg = disruption)

last_disruption_value <- tail(disruption, 1)
future_disruption <- c(disruption, last_disruption_value * exp(-0.05 * (1:456)))

```

###Forecast Using the Best SARIMAX Model 2 years for 19LGAs (for Cross River State) by 12months = 456 datapoints

```{r}
future_disruption <- future_disruption[1:456]  # Restrict future periods to 456

# Step 3: 
final_forecast <- forecast(cases_fit, xreg = future_disruption, h = 456)
final_forecast
```

##Plot historical values and forecast while maintaining correct limits

```{r}
autoplot(dataMIP_ts) + 
  autolayer(final_forecast, series = "Forecast", PI = TRUE) +
  ggtitle("SARIMAX(2,0,2)(0,0,0) Forecast for CrossRiver State") +
  xlab("Time (months)") + 
  ylab("MIP Cases") +
  scale_x_continuous(breaks = seq(2015, 2027, 1)) +
  theme_minimal()

```
